Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/algorithms.tex
edits to evaluation to make it logically more consistent with the study

\noindent
Checking verifies the access pattern of an array computation in the
source language against any associated specifications. Checking
proceeds by generating a model from a specification and generating a
model from the source code (above), and comparing them for
consistency.  Since the model of \Cref{sec:model} interprets both
array indices and specifications as sets of points in $N$-dimensional
space. The notion of consistency is then intuitively whether these two
(potentially infinite) sets of points are equal. This leads to a
simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
tests the consistency of a modal $M_C$ of source code against
a model $M_S$ of a specification:
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked by the \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

@@

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, the performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times the number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment. In \Cref{subsec:additional-data}, we
established that  99.6\% of array computations have 
dimensionality of at most four, and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show the number of regions that are
connected with no more than two \texttt{+} operations in 100\% of the cases.

@@

  \item Interval schemes into a joins of meets
  of region constants, where some intervals are split into separate
  region constants.
\end{enumerate}
%
An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
each of vector can be represented as a vector of holed intervals
$\interv{a}{b}{c}$.  If so, we convert each index scheme to an
interval scheme as a vector of holed interval representation.

Since index schemes of the previous stage are all contiguous and form
closed intervals, a lower bound $\leq 0$ and an upper bound $\geq 0$
is sufficient to form an interval scheme.  Otherwise, an approximate
specification is generated.

@@

Recall the model of region constants in \Cref{subfig:region-model}
where $\interp{-}$ maps to holed intervals which have either $0$ as
the lower-bound or upper-bound for \fortran{forward} and
\fortran{backward} respectively, or both $0$ for \fortran{pointed}, or
$-k$ and $k$ lower and upper bounds for \fortran{centered}. We can
invert this map to generate region constants from holed
intervals. However, some intervals computed from the above steps
may not the constraints of this interpretation,
\eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
\Cref{lem:vector-union}, we can split interval schemes into two into
the relevant form, for example, producing
$\interv{-2}{0}{} \times \interv{-1}{1}{}$ and
$\interv{0}{1}{} \times \interv{-1}{1}{}$. 
The following function
$\textit{split}$ iterates over the components of an index scheme,
splitting apart interval schemes when necessary:
%
\begin{align*}
  %\textit{split}(\emptyset) = \emptyset \quad
  \textit{split}(\interv{a}{b}{c} \times S) =
   \begin{cases}
  \{ \interv{a}{0}{c} \times T \mid T \in \textit{split}(S) \}
  \cup
  \{ \interv{0}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & a < 0 \vee b > 0 \vee (|a| \neq b) \\
  \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
\end{cases}
\end{align*}
%
This forms a set of interval schemes to which $\interp{-}^{-1}$ can be
applied. In the case of our example, no splitting is necessary.

The final stage of inference is simply mapping interval schemes into
region constants and combining these. Each translated interval scheme
is connected with \texttt{+}. Individual schemes are decomposed using
\Cref{lem:vector-intersect} such that each of the resulting index
schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.

Each of these interval schemes are mapped to region constants using
\Cref{subfig:region-model} and connected with \texttt{*}.
%
\begin{align*}
\textit{convert}_N(\mathcal{S}) = \sum_{S \in S} \prod_{1 \leq i \leq N} \interp{\pi_i(S)}^{-1}
\end{align*}
%
In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. The decomposition of
$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
following specification:
%
\begin{equation*}
  \texttt{stencil readOnce,} \;
  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
\end{equation*}

>>>>

\noindent
Checking verifies the access pattern of an array computation in the
source language against any associated specifications. Checking
proceeds by generating a model from a specification and generating a
model from the source code (above), and comparing them for
consistency.  Since the model of \Cref{sec:model} interprets both
array indices and specifications as sets of points in $N$-dimensional
space. The notion of consistency is then intuitively whether these two
(potentially infinite) sets of points are equal. This leads to a
simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
tests the consistency of a modal $M_C$ of source code against
a model $M_S$ of a specification:
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The $\mathit{consistent}$ function checks whether linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. It then delegates to $\mathit{consistent'}$
to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked \texttt{atLeast},
require the space defined by the specification to remain inside the
set of indices, while an upper bound, marked by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

@@

Although satisfiability is super-exponential in the length of the formula in
the worst case~\cite{fischer1974super}, the performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times the number of regions composed
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment. In \Cref{subsec:additional-data}, we
established that 99.6\% of array computations have 
dimensionality at most four and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show that all specifications in the corpus
comprises no more than two \texttt{+} operations. Thus, our approach
using Z3 is practical and efficient in all but corner cases.

@@

  \item Interval schemes into a joins of meets
  of region constants, where some intervals are split into separate
  region constants.
\end{enumerate}
%
An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
each vector can be represented as a vector of holed intervals
$\interv{a}{b}{c}$.  Since index schemes of the previous stage are all
contiguous and form closed intervals, a lower bound $\leq 0$ and an
upper bound $\geq 0$ is sufficient to form an interval scheme.
Otherwise, an approximate specification is generated.

@@

Recall the model of region constants in \Cref{subfig:region-model}
where $\interp{-}$ maps to holed intervals which have either $0$ as
the lower-bound or upper-bound for \fortran{forward} and
\fortran{backward} respectively, or both $0$ for \fortran{pointed}, or
$-k$ and $k$ lower and upper bounds for \fortran{centered}. We can
invert this map to generate region constants from holed
intervals. However, some intervals computed from the above steps
may not the constraints of this interpretation,
\eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
\Cref{lem:vector-union}, we can split interval schemes into two into
the relevant form, for example, producing
$\interv{-2}{0}{} \times \interv{-1}{1}{}$ and
$\interv{0}{1}{} \times \interv{-1}{1}{}$. 
The following function
$\textit{split}$ iterates over the components of an index scheme,
splitting apart interval schemes when necessary:
%
\begin{align*}
  %\textit{split}(\emptyset) = \emptyset \quad
  \textit{split}(\interv{a}{b}{c} \times S) =
   \begin{cases}
  \{ \interv{a}{0}{c} \times T \mid T \in \textit{split}(S) \}
  \cup
  \{ \interv{0}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & a < 0 \vee b > 0 \vee (|a| \neq b) \\
  \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
\end{cases}
\end{align*}
%
This forms a set of interval schemes to which $\interp{-}^{-1}$
(\Cref{subfig:region-model}) is well-defined for each component.

The final stage of inference, thus maps the component
of every interval scheme to a region constant via $\interp{-}^{-1}$,
combining each component with \texttt{*} can combining the
resulting regions by \texttt{+}, that is:
%
%interval schemes to region constants and combines them. %Individual
%schemes are decomposed using \Cref{lem:vector-intersect} such that
%each of the resulting index schemes have all but one dimensions set to
%\interv{-\infty}{\infty}{}.
%Each of these interval scheme are mapped to region constants using
%\Cref{subfig:region-model} and connected with \texttt{*}.
%
\begin{align}
\textit{convert}_N(\mathcal{S}) = \sum_{S \in \mathcal{S}} \; \prod_{\pi_i(S) \neq
  \interv{-\infty}{\infty}{}} \interp{\pi_i(S)}^{-1}
  \label{eq:convert}
\end{align}
%
Where summation is by the syntactic \term{+} and product is by the
syntactic \term{*}

In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is mapped to
\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}, and
$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined by
\term{+} we obtain the following specification:
%
\begin{equation*}
  \texttt{stencil readOnce,} \;
  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
\end{equation*}
\vspace{-1em}
\begin{restatable}[Inference soundness]{lem}{inferenceSoundness}
  For all region specs $R$, then $\textit{infer}(\interp{R}_N)_N \equiv R$
\end{restatable}

### 2017-04-18 oopsla17/evaluation.tex
edits to evaluation to make it logically more consistent with the study

To study the effectiveness of our approach, we applied our tool to the
more than one million lines of source code contained within the corpus
that we have assembled (\Cref{sec:study}). We first examined how
frequently stencil computations occur. We parsed
\num{\overalllinesParsed} lines of Fortran code and found that
\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
have a left-hand side as an array subscript on neighbourhood
indices. This supports the idea that stencil-like computations are
common in scientific code.  We then used the inference procedure of
the previous section to generate specifications for stencils in the
corpus to assess the design of the language.

We would not expect to infer a stencil for each array statement we
found because our analysis restricts the array-subscript-statements
that we classify as a stencil. For example, \num{\overalltickLHSvar}
potential stencils were discarded because the left-hand side was not
an array access. Overall, we were able to infer a stencil from
\overalltickAssignSuccessPercentOfTickAssign\% of the statements
involving an array assignment. A single statement can involve multiple
arrays and we ended up with \num{\overallnumStencilSpecs}
specifications. This shows that we can express a large number of
stencil shapes within our high-level abstraction and validates our
initial hypothesis that many stencil computations have a regular
shape.

@@

\textbf{All pointed} \num{\overalljustPointed} of the stencils we
found involved only \pointed\ regions.
% 39,681 of these were pointed in all dimensions.
Common examples of this were pointwise transformations on data (such
as scaling).

@@

We measured the frequencies of occurrences of the operators \term{*} and \term{+} within stencil specifications inferred from our corpus:
\begin{center}
\vspace{0.25em}
\setlength{\tabcolsep}{0.57em}
{\small{
\hspace{-1em}\begin{tabular}{c|cccccccc}
Occurrences: & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
\term{*} & \num{\overallmulOpszero} & \num{\overallmulOpsone} & \num{\overallmulOpstwo} & \num{\overallmulOpsthree} & \num{\overallmulOpsfour} &  & \num{\overallmulOpssix} \\
\term{+} & \num{\overallplusOpszero} & \num{\overallplusOpsone} & \num{\overallplusOpstwo}
\end{tabular}}}
\vspace{0.2em}
\end{center}

@@

There were various reasons why we did not infer
specifications on every looped array computation:
\begin{itemize}
\item \textbf{Derived induction variables} where an
index (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in
\mintinline{fortran}{x = len - i};

\item \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
\end{itemize}

\subsection{Detecting errors in the 2-D Jacobi iteration}
\label{sec:jacobi}
One common example of a stencil computation is the two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
is given by:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
We infer a precise specification of its shape as:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)
\end{minted}
%
To test our implementation,
we examined whether programmer errors would be detected by replacing
the array index offsets with $-1$, $0$, or $1$ and running our
verification algorithm. Our checking procedure correctly reported a verification
failure in each of 6,537 permutations corresponding to an error.
The iteration computes the average of four adjacent
cells so $24$ ($4$ factorial) of the possible array index perturbations
are correct, all of which are accepted by our checker.

@@

\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil readOnce, forward(depth=1, dim=3, nonpointed)*pointed(dim=1)*pointed(dim=2)
\end{minted}

>>>>

\noindent
To study the effectiveness of our approach, we applied our tool back
onto the corpus from \Cref{sec:study}.  We used the inference
procedure of the previous section to generate array specifications
for the corpus to assess the design of the language.
%
%We first examined how
%frequently stencil computations occur. We parsed
%\num{\overalllinesParsed} lines of Fortran code and found that
%\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
%have a left-hand side as an array subscript on neighbourhood
%indices. This supports the idea that stencil-like computations are
%common in scientific code.  
%
We did not expect to infer specifications for each array computation
we found because our analysis restricts the
array-subscript-statements, along the lines of~\Cref{sec:study}.
%A
%single statement can involve multiple arrays and therefore can have
%multiple specifications associated with it.
Overall we ended up with
\num{\overallnumStencilSpecs} specifications of which
\num{\overalltickLHSvar} were with a left-hand side variable rather
than a subscript (an \term{access} specification), leaving
\num{\overallStencilsSpecs} \term{stencil} specifications.
%
%For example, \num{\overalltickLHSvar}
%potential stencils were discarded because the left-hand side was not
%an array access. Overall, we were able to infer a stencil from
%\overalltickAssignSuccessPercentOfTickAssign\% of the statements
%involving an array assignment. A single statement can involve multiple
%arrays and.
This shows that we can express a large number of
array access patterns within our high-level abstractions and validates our
initial design choices informed by our data (~\Cref{subsec:choices}).

@@

\textbf{All pointed} \num{\overalljustPointed} of the specifications we
generated involved only \pointed\ regions.
% 39,681 of these were pointed in all dimensions.
Common examples of this were pointwise transformations on data (such
as scaling).

@@

We measured the frequencies of occurrences of the operators \term{*}
and \term{+} within the inferred specifications:
\begin{center}
\vspace{0.25em}
\setlength{\tabcolsep}{0.57em}
{\small{
\hspace{-1em}\begin{tabular}{c|cccccccc}
Occurrences: & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
\term{*} & \num{\overallmulOpszero} & \num{\overallmulOpsone} & \num{\overallmulOpstwo} & \num{\overallmulOpsthree} & \num{\overallmulOpsfour} &  & \num{\overallmulOpssix} \\
\term{+} & \num{\overallplusOpszero} & \num{\overallplusOpsone} & \num{\overallplusOpstwo}
\end{tabular}}}
\vspace{0.2em}
\end{center}

@@

\noindent
There were various reasons why we did not infer
specifications on every looped array computation:
\begin{itemize}
\item \textbf{Computed induction variables} where an
index (\mintinline{fortran}{x}) is computed from an
induction variable via an intermediate definition, \eg{}, 
\mintinline{fortran}{x = i + 1}, or using functions that
we could analyse or classify;

\item \textbf{Inconsistent LHS/RHS} occur when
induction variable is used to specify more than one array dimension
on the RHS or multiple induction variables are used for the same
dimension on the RHS. These are common in matrix operations such as
LU-decomposition with assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
\end{itemize}

\subsection{Detecting errors in the 2-D Jacobi iteration}
\label{sec:jacobi}

\noindent
One common example of a stencil computation is the two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
is given by:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
We infer a precise specification of its shape as:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)
\end{minted}
%
To test our implementation,
we examined whether programmer errors would be detected by replacing
the array index offsets with $-1$, $0$, or $1$ and running our
verification algorithm. Our checking procedure correctly reported a verification
failure in each of 6,537 permutations corresponding to an error.
The iteration computes the average of four adjacent
cells so $24$ ($4$ factorial) of the possible array index perturbations
are correct, all of which are accepted by our checker.

@@

\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil readOnce, forward(depth=1, dim=3, nonpointed)*pointed(dim=1)*pointed(dim=2)
\end{minted}
%
We found a similar bug in 
Cliffs\footnote{\url{https://github.com/Delta-function/cliffs-src/commit/e2ca312cfc5398287c7ca8594e977658617c2540}}
where a subscript \fortran{u(i)} is corrected to \fortran{u(i-1)} which
could have been detected by an initial specification \fortran{backward(depth = 1,
  dim = 1, nonpointed)}.

### 2017-04-18 oopsla17/proofs.tex
edits to evaluation to make it logically more consistent with the study

>>>>

\inferenceSoundness*

\begin{proof}
  This follows by induction on the definition of regions syntax
  $R$. and by definitions of inference and the semantics sharing the
  same underlying model and using inverse mappings between each other.

  For constants, we map to
  $\vecgen{N}{i}{\interp{\texttt{\textcap{rconst}}}}$
  which by \textit{convert}, in \eqref{eq:convert} is mapped
  back to $\textcap{rconst}$ since \textit{convert} is defined
  in terms of the inverse mapping $\interp{-}^{-1}$.

  For $\interp{r \term{+} s}_N = \interp{r}_N \vee \interp{s}$
  we map to a Union Normal Form of interval spaces which is convert
  back directly by \textit{convert}. Similarly for $\term{*}$.
\end{proof}
